---
title: "RWorkshop"
subtitle: "Linear regression"
author: "PT"
date: "Aug 2021"
output: ioslides_presentation
---

```{css, echo=FALSE}
pre {
  max-height: 400px;
  overflow-y: auto;
}

pre[class] {
  max-height: 100px;
}
```

```{r, echo=FALSE, message=FALSE, warning=FALSE}
library(ggplot2)
```


# Linear Regression


---

Estimate the regression coefficents in
$$
\hat{testscr} = \hat{\beta_0} + \hat{\beta_1}str
$$
Where, $testscr = (math + read)/2$ is the average test score in the district, and $str = students/teachers$ is the student-teacher ratio.

- The data set `CASchools` is in the `AER` package.
1. Install the `AER` package.
1. Load the `CASchools` dataset.
1. Create new variables $testscr$ and $str$.
1. Estimate $\hat\beta_0$ and $\hat\beta_1$.


---

```{r, warning=FALSE, message=FALSE}
library(AER)
data(CASchools, package = "AER")
caschool <- CASchools
caschool$str <- caschool$students/caschool$teachers
caschool$testscr <- (caschool$math + caschool$read)/2
```

```{r}
names(CASchools)
```
```{r}
names(caschool)
```

---


```{r,  fig.cap="Scatterplot of TestScore vs Student-Teacher ratio (California School District Data"}
plot(caschool$testscr ~ caschool$str,
     xlab = "Student-teacher ratio",
     ylab = "Test score", pch = 20, xlim = c(10,30),
     ylim = c(600, 720), las = 1)
```


---

```{r}
par(mfrow = c(1,2))
hist(caschool$testscr, main = "Test Scores",
     xlab = "testcr", ylab = "Frequency")
hist(caschool$str, main = "Student Teacher Ratio",
     xlab = "str")


```

---

```{r, message=FALSE, warning=FALSE}
summarytools::dfSummary(caschool[, c('testscr', 'str')], graph.col = FALSE)
# summary() can also be used.
```


---



- The `lm()` function is used to fit a <u>**L**</u>inear <u>**M**</u>odel.
- Usage
```{r, eval = FALSE}
lm(formula, data, subset)
```
- The California Schools model
```{r}
# fit a linear model and save it to an `lm` object
lm_model1 <- lm(formula = testscr ~ str, data = caschool)
class(lm_model1)
```

---

You can directly print the `lm_model1` output
```{r}
lm_model1
```


---

Or, you can also print the `summary()`.
```{r}
summary(lm_model1)
```

---

- Add a regression line.
- First, we shall extract the coefficients from `lm_model1`
```{r}
b1 <- coef(lm_model1)[[1]]
b2 <- abs(coef(lm_model1)[[2]])

```

- We shall also need `expr()` from `rlang` package to add Mathematical Annotation in R.

```{r, message=FALSE, warning=FALSE}
library(rlang) 
```
- Use `?plotmath()` and `demo(plotmath)` for usage.


## {.smaller .vcenter}


```{r}
plot(caschool$testscr ~ caschool$str,
     xlab = "Student-teacher ratio",
     ylab = "Test score", pch = 20, xlim = c(10,30),
     ylim = c(600, 720), las = 1)

# add the model fit
abline(lm_model1, col = "blue", lwd = 2)
text(25, 710, expr(paste(widehat(TestScore)~"="~ !!b1~"-"~ !!b2~X~STR)), cex = 0.8)
arrows(25,702, 28,640, length = 0)
```



# The omitted variable(s)

## {.smaller}
- What was our previous assessment about `testscr`?
    - We estimated the linear model $\hat{testscr} = 698.9-2.27\times str$
   - Interpretation: Class with higher `str` i.e. larger class size tend to have lower `testscr`.

- Reasons for considering `english` as omitted variable;
    - Large immigration population in California
    - Students in the district who are still learning English.
    - Students who are still learning English might perform worse on standardized tests as compared to native English speakers. 
    - The larger class sizes `str` might also have students still learning English. Therefore, considering the OLS estimate $\hat{testscr} = 698.9-2.27\times str$, the policy would be considering lowering the `str` value to improve `testscr`. But this might not solve the problem because of the English learning students in the class.


## {.smaller}

- Our multiple regression model is
$$
\hat{testscr} = \beta_0 + \beta_1str + \beta_2english
$$
where $english$ is the percentage of students in the district who are English learners.
```{r}
# summary stats for the three variables
summary(caschool[c("testscr", "str", "english")])
```

---

```{r}
summarytools::descr(caschool[c("testscr", "str", "english")])
```

---

```{r}
summarytools::dfSummary(caschool, graph.col = FALSE)
```


## {.smaller}

**Estimated model**

```{r}
lm_multi <- lm(testscr ~ str + english, data = caschool)
summary(lm_multi)
```

# Regression when $X$ is a binary variable

## {.smaller}

- **Equation with a binary regressor**: 
$$Y_i = \beta_0 + \beta_1 D_i + u_i$$
$$D_i = \begin{cases}
        1 \ \ \text{if $STR$ in $i^{th}$ school district < 20} \\
        0 \ \ \text{if $STR$ in $i^{th}$ school district $\geq$ 20} \\
      \end{cases}$$
- Our regression model is
$$
testscr_i = \beta_0 + \beta_1 D_i + u_i.
$$

**What is the difference in the expected test scores in districts with lower student-teacher ratio $STR < 20 (D_i = 1)$ and those with higher student-teacher ratio $STR \geq 20 (D_i=0)$?**

## {.smaller}

- Create the dummy variable and append it to the dataset.
```{r}
# create D and append
caschool$D <- as.factor(ifelse(caschool$str < 20, 1, 0))

# check the dataset
str(caschool)
```

## {.smaller}
- Plot the data
```{r}
plot(testscr ~ D, data = caschool, main = "Dummy Regression")
legend("topleft", legend = c(": STR < 20", expression(paste(": ",STR >= 20))),
       pch = c("1", "0"), cex = 0.7)
```


## {.smaller}

**Estimate the dummy regression model**
```{r}
lm_dummy <- lm(testscr ~ D, data = caschool)
summary(lm_dummy)
```

## {.smaller}

**The interpretation of the coefficients in this regression model is as follows:**

$\beta_0$ = `r round(coef(lm_dummy)[[1]], 2)` is the expected test score in districts where $STR \geq 20$.

$\beta_1$ = `r round(coef(lm_dummy)[[2]], 2)` is the difference in the expected test score between districts with $STR < 20$ and those with $STR \geq 20$.
    
**Observation:**

One can see that the expected test score in districts with STR<20 ($D_i=1$) is predicted to be 650.1+7.17=657.27 while districts with STR≥20 ($D_i=0$) are expected to have an average test score of only 650.1.

# Multiple models in `stargazer` with heteroskedasticity-robust standard errors.

---

We shall consider the following models and graphically inspect the correlation and summarise all the regression results in a table.
$$
\begin{align*}
\text{(I)  }\text{testscr} &= \beta_0 + \beta_1\text{str} + u\\
\text{(II)  }\text{testscr} &= \beta_0 + \beta_1\text{str} + \beta_2\text{english} + u\\
\text{(III)  }\text{testscr} &= \beta_0 + \beta_1\text{str} + \beta_2\text{english} + \beta_3\text{lunch} + u\\
\text{(IV)  }\text{testscr} &= \beta_0 + \beta_1\text{str} + \beta_2\text{english} + \beta_3\text{calworks} + u\\
\text{(V)  }\text{testscr} &= \beta_0 + \beta_1\text{str} + \beta_2\text{english} + \beta_3\text{lunch} + \beta_4\text{calworks} + u\\
\end{align*}
$$

where, `testscr`- test score
`str` - student-teacher ratio
`english` - percentage of english learners
`lunch` - percentage of students that qualify for a free or subsidised lunch due to family incomes below a certain threshold.
`calworks` - percentage of students qualifying for the *calworks* income assistance programme.

---

```{r}
par(mfrow = c(2,2))
plot(testscr ~ str, data = caschool, main = "Test Scores and\nStudent teacher ratio")

plot(testscr ~ english, data = caschool, main = "Test Scores and\nPercentage of english learners")

plot(testscr ~ lunch, data = caschool, main = "Test Scores and\nPercentage of students qualifying\nfor free lunch")

plot(testscr ~ calworks, data = caschool, main = "Test Scores and\nPercentage of students qualifying \nfor income assistance programme")
```


## {.smaller}

```{r,  warning=FALSE}
# models
I <- lm(testscr ~ str, data = caschool)
II <- lm(testscr ~ str + english, data = caschool)
III <- lm(testscr ~ str + english + lunch, data = caschool)
IV <- lm(testscr ~ str + english + calworks, data = caschool)
V <- lm(testscr ~ str + english + lunch + calworks, data = caschool)

# gather all the robust standard error values in a list
robust_se <- list(
  sqrt(diag(sandwich::vcovHC(I, type = "HC1"))),
  sqrt(diag(sandwich::vcovHC(II, type = "HC1"))),
  sqrt(diag(sandwich::vcovHC(III, type = "HC1"))),
  sqrt(diag(sandwich::vcovHC(IV, type = "HC1"))),
  sqrt(diag(sandwich::vcovHC(V, type = "HC1")))
)

## use the `stargazer()` function to represent the output in a tabular form
stargazer::stargazer(I, II, III, IV, V, type = "text", 
                     title = "Analysis of caschool dataset.",
                     se = robust_se, digits = 3, 
                     column.labels = c("I", "II", "III", "IV", "V"),
                     column.sep.width = "1pt",
                     notes = "Heteroskedasticity robust standard errors are given in parentheses under coefficients.", 
                     notes.append = TRUE) 
```
